---
title: "Citations for R and packages used"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis', warning = F)
```

The authors of this project would like to thank the countless hours of work that have made our analysis possible. In addition to the curators of the [COSMIC and CGC](https://doi.org/10.1093/nar/gkw1121), and the [UCSC genome browser](https://genome.ucsc.edu), we thank the authors of the following free and open-source software pacakges.

The R project

```{r}
print(citation(), style = 'html')
```

[The RStudio IDE](https://www.rstudio.com)

R packages to streamline data analysis

```{r}
print(citation('tidyverse'), style = 'html')
print(citation('dplyr'), style = 'html')
print(citation('tidyr'), style = 'html')
print(citation('purrr'), style = 'html')
print(citation('readr'), style = 'html')
print(citation('stringr'), style = 'html')
print(citation('tibble'), style = 'html')
print(citation('fuzzyjoin'), style = 'html')
```

R packages for genomic sequence analysis

```{r}
print(citation('BSgenome'), style = 'html')
print(citation('Biostrings'), style = 'html')
print(citation('IRanges'), style = 'html')
print(citation('GenomicRanges'), style = 'html')
print(citation('biomaRt'), style = 'html')
```

R packages for figure generation

```{r}
print(citation('ggplot2'), style = 'html')
```
