---
title: "iSTOP Paper"
output:
  md_document:
    variant: markdown_github
    toc: yes
  word_document:
    toc: no
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This project accompanies the [iSTOP](github.com/CicciaLab/iSTOP) R package, and is intended to aid reproduction of figures for a forthcoming publication detailing the scope of possible nonsense mutations capable using [CRISPR mediated base editing](https://www.nature.com/nature/journal/v533/n7603/abs/nature17946.html). If you just want to search for iSTOP targetable sites in your favorite gene, you should head over the the [iSTOP](github.com/CicciaLab/iSTOP) R package page.



# Installation

To reproduce the analysis in [TBD](link/to/publication), begin by installing [R](https://cran.r-project.org) (~100 MB) and [RStudio](https://www.rstudio.com/products/rstudio/download/) (~500 MB). 

Then, clone this project to your computer using the big green `Clone or download>Download ZIP` button near the top of [this page](github.com/CicciaLab/iSTOP). Then unzip and open the project by clicking on the `iSTOP-paper.Rproj` file. This will open RStudio with the working directory set to the project's folder.

## System requirements

This analysis requires ~11 GB of disk space, at least 4 GB of memory, and a little bit of patience. While the analysis should be cross-platform, it has only been tested on macOS Sierra Version 10.12.4 with a 4 GHz processor and 32 GB of memory. Processing time on this system using 4 cores for just the Human genome is ~30 minutes to locate all iSTOP codons and targets, and an additional ~40 minutes for each RFLP annotation width (i.e. `add_RFLP(width = 150)` takes ~40 minutes).

## R packages

To install all necessary R packages, run the following R commands in the RStudio console.

```{r}
# Source the Biocoductor installation tool - installs and loads the 
# BiocInstaller package which provides the biocLite function.
source("https://bioconductor.org/biocLite.R")

# Install the following required packages ~ 350 MB 
BiocInstaller::biocLite(c(
  # Packages from CRAN (cran.r-project.org)
  'tidyverse',
  'gridExtra',
  'cowplot',
  'assertthat',
  'pbapply',
  'devtools',
  # Packages from GitHub (github.com) (version with genome_*_join)
  'dgrtwo/fuzzyjoin',
  # Packages from Bioconductor (bioconductor.org)
  'BSgenome',
  'Biostrings',
  'GenomicRanges',
  'IRanges'
))

# Install genomes and annotation packages ~ 3 GB
BiocInstaller::biocLite(c(
  # Arabidopsis (not available from UCSC)
  'TxDb.Athaliana.BioMart.plantsmart28', # ~ 24  MB
  'org.At.tair.db',                      # ~ 239 MB
  # Genomes
  'BSgenome.Hsapiens.UCSC.hg38',         # ~ 802 MB
  'BSgenome.Celegans.UCSC.ce11',         # ~ 25  MB
  'BSgenome.Dmelanogaster.UCSC.dm6',     # ~ 36  MB
  'BSgenome.Drerio.UCSC.danRer10',       # ~ 343 MB
  'BSgenome.Mmusculus.UCSC.mm10',        # ~ 683 MB
  'BSgenome.Rnorvegicus.UCSC.rn6',       # ~ 719 MB
  'BSgenome.Scerevisiae.UCSC.sacCer3',   # ~ 3   MB
  'BSgenome.Athaliana.TAIR.TAIR9'        # ~ 34  MB
))

# If all goes well, install the iSTOP package hosted on GitHub
BiocInstaller::biocLite('CicciaLab/iSTOP')
```

## Datasets

### COSMIC

Download the "COSMIC Mutation Data" (~300 MB compressed) from the [Catalogue of Somatic Mutations in Cancer](https://cancer.sanger.ac.uk/cosmic/download). This dataset requires [registration with a valid email address](https://cancer.sanger.ac.uk/cosmic/register), then in your terminal (not the R console!) you can download the file with the following commands (substitute `your_email_address` with the email used to register).

```
sftp "your_email_address"@sftp-cancer.sanger.ac.uk
# You will be prompted for the password you provided when you registered
get /files/grch38/cosmic/v80/CosmicMutantExport.tsv.gz
```

Move this file to the `data/COSMIC` directory of this project. It should already be named `CosmicMutantExport.tsv.gz`.

### CGC

Download the [Cancer Gene Census (CGC) dataset](http://cancer.sanger.ac.uk/census) by clicking on the `CSV` Export button. You will need to login with the same credentials used to download the COSMIC dataset. Move this file to the `data/COSMIC` directory of this project. Rename the file `CGC.csv`.

### CDS

Download CDS coordinates for each genome (~120 MB). Back in the RStudio console, run the following commands.

```{r}
library(tidyverse)
library(iSTOP)

CDS_Athaliana_BioMart_plantsmart28() %>% 
  write_csv('data/CDS/Athaliana-plantsmart28.csv')
CDS_Celegans_UCSC_ce11() %>% 
  write_csv('data/CDS/Celegans-ce11.csv')
CDS_Dmelanogaster_UCSC_dm6() %>% 
  write_csv('data/CDS/Dmelanogaster-dm6.csv')
CDS_Drerio_UCSC_danRer10() %>% 
  write_csv('data/CDS/Drerio-danRer10.csv')
CDS_Hsapiens_UCSC_hg38() %>% 
  write_csv('data/CDS/Hsapiens-hg38.csv')
CDS_Mmusculus_UCSC_mm10() %>% 
  write_csv('data/CDS/Mmusculus-mm10.csv')
CDS_Rnorvegicus_UCSC_rn6() %>% 
  write_csv('data/CDS/Rnorvegicus-rn6.csv')
CDS_Scerevisiae_UCSC_sacCer3() %>% 
  write_csv('data/CDS/Scerevisiae-sacCer3.csv')
```

# Comprehensive search for iSTOP targetable sites

Load all required packages and functions with the following commands.

```{r}
library(tidyverse)
library(stringr)
library(iSTOP)

# Source all R functions defined in this project
list.files('R/functions', '[.]R', full.names = T) %>% walk(source)
```

Given CDS coordinates and genomes, search for all iSTOP sites with the following commands. Raw results will be saved to the `data/iSTOP` directory, and a compacted version with RFLP annotations will be saved to the `data/iSTOP-compact` directory. To dramatically reduce computation time, comment out the `add_RFLP` lines. The `add_RFLP` lines for the Human dataset are only required to reproduce Figure 3F and Supplementary Figure 3A.

```{r}
# Adjust the number of cores for parallel computation to suit your computer
# Assume that each core will require ~2.5 GB of Memory
# Set to 1 or 2 if you are unsure. Only 1 core is supported on Windows
cores = 1

# Only the human datasets are necessary to reproduce Figure 3
read_csv('data/CDS/Hsapiens-hg38.csv', col_types = 'cciccii') %>%
  locate_codons(BSgenome.Hsapiens.UCSC.hg38::Hsapiens, cores = cores) %>%
  locate_iSTOP(BSgenome.Hsapiens.UCSC.hg38::Hsapiens) %>%
  write_csv('data/iSTOP/Hsapiens-hg38.csv') %>%       # ~ 1 GB
  compact_iSTOP %>%
  add_RFLP(width = 150, cores = cores) %>%
  add_RFLP(width = 100, cores = cores) %>%
  add_RFLP(width = 50,  cores = cores) %>%
  write_csv('data/iSTOP-compact/Hsapiens-hg38.csv')

# The remaining species datasets are necessary to reproduce Figure 4
# This script will run the above command for the non-human species
source('R/scripts/iSTOP-non-human-species.R')
```

Once complete, compute summaries by codon and untargetable datasets with the following command.

```{r}
# Writes two files each to 
# `data/iSTOP-by-codon` and `data/iSTOP-untargetable`
list.files('data/iSTOP', '[.]csv', full.names = T) %>% 
  pbapply::pblapply(summarize_by_codon, cl = cores)

# Summarize targetability for all species on codon, ORF and gene levels
source('R/scripts/summarize-by-codon-ORF-gene.R')
```

# Analysis of COSMIC nonsense mutations

The raw COSMIC dataset can be cleaned and summarized by sourcing the `R/Clean-COSMIC.R` script. This will add three datasets to the `data/COSMIC` directory.

1. `COSMIC-iSTOP.csv` - (~360 MB) All frameshift and substitution mutations for GRCh38, with aggregated cancer types, and annotated as to whether or not the mutation corresponds to an iSTOP targetable coordinate.
2. `COSMIC-nonsense.csv` - (~7 MB) Only nonsense mutations from `COSMIC-iSTOP.csv`
3. `COSMIC-summary-by-cancer.csv` - Summary by cancer type that details the frequency of nonsense, and targetability with iSTOP
4. `COSMIC-summary-by-gene.csv` - Summary by gene that includes test results for frequent stoppers (likely tumor suppressors). Note that test results for "All cancers" are simply the smallest observed p-value across all cancer subtypes for a given gene.

```{r}
source('Clean-COSMIC.R')
```

# Reproducing Figures

Each script will write figures as PDF files to the `figures` directory of the project. Final figures for the paper were edited in [Inkscape](https://inkscape.org/en/) to reduce the size of the files and improve readability of figure legends and axis labels.

```{r}
library(tidyverse)
library(iSTOP)
library(stringr)
library(gtable)
library(gridExtra)
library(cowplot)

source('R/figures/Figure-1.R')
source('R/figures/Figure-3.R')
source('R/figures/Figure-4.R')
source('R/figures/Figure-5.R')
source('R/figures/Supp-Figure-3.R')
source('R/figures/Supp-Figure-5.R')
```

Figure 5E demonstrates the utility of the iSTOP package for visualizing iSTOP targetable sites. This figure can be reproduced for any gene using the following commands.

```{r}
library(tidyverse)
library(iSTOP)

my_gene = 'ATM'

COSMIC <- read_csv('data/COSMIC/COSMIC-nonsense.csv')
CDS    <- read_csv('data/CDS/Hsapiens-hg38.csv')

iSTOP <-
  filter(CDS, gene == my_gene) %>%
  locate_codons(BSgenome.Hsapiens.UCSC.hg38::Hsapiens) %>%
  locate_iSTOP(BSgenome.Hsapiens.UCSC.hg38::Hsapiens) %>%
  add_RFLP(width = 50)

Fig5E <- plot_spliced_isoforms(
  gene   = my_gene,
  coords = filter(CDS, tx %in% iSTOP$tx),
  colors = c('red', 'black', 'blue', 'darkgreen'),
  `Nonsense in cancer`   = COSMIC,
  `CAA, CAG, CGA, TGG`   = iSTOP,
  `iSTOP targetable`     = filter(iSTOP, match_any),
  `Verifiable with RFLP` = filter(iSTOP, match_any & has(RFLP_50))
)

Fig5E
ggsave('figures/Figure-5E.pdf', Fig5E, width = 18, height = 3)
```

```{r}
# Clean up workspace - Leave Figures and Figure data
rm(list = setdiff(ls(), ls(pattern = '^Fig|COSMIC|CDS|iSTOP')))
```

# Session Information

This analysis was successfully performed with the following system, and package versions:

```{r echo = F, eval = T}
library(tidyverse)
library(iSTOP)
library(stringr)
library(gtable)
library(gridExtra)
library(cowplot)
options(width = 90)
devtools::session_info()
```
